\startchapter{Glossary}
\label{chapter:glossary}

ANOVA (Science) - Analysis of Variance between groups - A group of statistical techniques to test hypotheses based on experimental results.  It tests the distribution of the means of variables to see if they are the same, assuming the variables are normally distributed.
\\

Cross-validation (Computer Science) - A method used to test the accuracy of machine learning classifiers by separating a dataset into examples used to train the classifier and examples used to test the classifier.
\\

Ethnomusicology (Musicology) - a sub-discipline of Musicology that focuses on the study of the socio-cultural aspects of music in societies around the world.
\\

Folksonomy (Computer Science) - a system of classification that comes from a group of users collaboratively creating and managing tags in an effort to annotate and categorize content
\\

Flux (MIR) - In this work, is the norm of the difference vector between two successive magnitue/power spectra.
\\

MFCC coefficients (MIR) - A way to transform a standard spectrum into one that more closely approximates how the human ear perceives sound.
\\
Music Information Retrieval (Computer Science) - also known by the acronym MIR, a new field of study where one applies tools from areas such as Digital Signal Processing, Audio Feature Extraction and Machine Learning to help people understand and retrieve information from music or audio.
\\

Ontology (Computer Science) - a formal representation of ideas or concepts and the relation between them.  This Computer Science definition of ontology is a subset of the broader philosophical idea of ontology, which is a study of what exists and what can exist.
\\

Rolloff (MIR) - A measure of the steepness of falloff in an audio spectrum
\\


Self-Organizing Map (Computer Science) - A technique that maps a high
dimensional feature space to a lower space.  It is a similar technique
to artifical neural networks.

Spectral Centroid (MIR) - A measure of the ``center of mass'' of a spectrum.
\\


XML (Computer Science) - eXtensible Markup Language - A popular
tree-tree based format for encoding data

JSON - Javascript Object Notation.  A modern data-exchange format,
used extensively as a way for servers to send data to web-based
clients.

Redis - A modern key-value object store designed to primarily hold all
it's data in memory to facilitate fast retrieval by clients.  

Syrinx - One of the structures in birds that is used to produce sound

Phonic Lips - Also known as ``museau de singe'' or ``monkey lips'' -
The mechanism by which odontocetes produce sound

Odontocete - A suborder of cetaceans of which one of the
distinguishing features is they have teeth, in contrast to the
Mysticeti, which have baleen

Mysticeti - A suborder of cetaceans of which one of the distinguishing
features is they use baleen to filter prey from the water, in contrast
to the Odontocetes, which have teeth
